Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2019-12-13 manuscript.tex
un poco de texto

\Abbreviations{CETI: Communicating Extraterrestrial Intelligence, DE:
Discrete Event, GHZ: Galactic Habitable Zone}

@@

Simulations are suitable tools to analyze systems that evolve with
time and involve randomness.
%
In general, simulations require less assumptions and simplifications
than theoretical approaches, and can be applied to systems where a
theoretical model can not be found.
%
Certain complex stochastic processes can be efficiently modeled with
the discrete--event (hereafter, DE) simulation approach.
%
A system described with the DE paradigm is characterized by a set of
actors and events, where actors interact causally through a series of
events on a time line and process these events in chronological order
\citep{ptolemaeus_system_2014, chung_simulation_2003,
ross_simulation_2012}.
%
We follow this approach to simulate the statistical properties of a
set of intelligent civilizations (CETIs) in the Galaxy that have the
ability of emitting and communicating messages at light speed.
%
Since this method to explore several models easily, it is well suited
for the particular case of the difusion of intelligence signals in the
Galaxy and its dependence on the statistical properties proposed for
the rise and fall of the communicating capacity in the galactic
intelligent civilizations.
%
The system is defined by a number of actors that represent CETIs and
appear at different instants in time, generating events that produce
meaningful changes in the variables that describe the system, i.e., in
the arrangement of CETIs and their network of causal contacts.
%
Assuming some simple hypotheses, the discrete events method can be
performed taking into account a small number of variables, which allow
to analyze the variation of the results in the model parameter space.

In what follows, we describe the experimental setup chosen to estimate
the probabilities of causal contacts and several derived quantities in
terms of three independent parameters, namely, the mean time span
between the appeareance of consecutive communicating civilizations
($\tau_a$), their mean lifetime ($\tau_s$), and the maximum distance a
signal can be detected by another civilization ($D_{max}$).
Intuitively, the shorter $\tau_a$ and the larger $\tau_s$ and
$D_{max}$, there is a greater probability of the existence of causal
contacts between pairs of intelligent civilizations in the Galaxy
(Table \ref{T_simu_hypotheses}).
We also need to propose theoretical distribution functions for both
the distribution of lifespans and the distribution of the number of
CETIs per unit time \citep{maccone_evolution_2014,
Sotos_biotechnology_2019}.
The shapes of these distributions are discussed in
Sec.~\ref{SS_PDF_shape}.

@@

involved civilizations being concurrent or active at the same time.

@@

Although a CETI could be old enough to transmit a message at large
distances, the limited power of the message and the dilution that runs
as the squared distance from the source, imposes a detectability
limit.

@@

communicating capacity, each CETI will fill a spherical shell region
of the Galaxy, limited by two surfaces.

@@

central CETI until it reaches the maximum distance $D_{max}$.

@@

Thus, the volume reached causally by a CETI is initially a growing

@@

the central CETI, with a delay with respect to the SFC equivalent to
the lifetime of the CETI, and produces a spherical shell region.

@@

Any other CETI within this region will be aware of the originating
CETI, even if it has disapeared before the time of contact.

@@

until the surface of last contact reaches $D_{max}$, and producing as
a result the loss of all signal from the central CETI.

@@

disk is negligigle with respect to the radius of the disk, so that
CETIs are placed on a 2D plane disk.

@@

of the region causally connected to the central CETI.

@@

At the end of the lifetime of a CETI, a growing surface of last

@@

In this scheme, the entire lifetime of a CETI can then be represented

@@

that result from the existence and communications of two CETIs (panel (b)).

@@

The points in time where the CETIs aquire their communicating
capacity (CC) are dubbed ''awakening'' events ($A_1$ and $A_2$ in the
Figure).

@@

Simmilarly, the points in time where the CETIs loss their communicating
capacity are dubbed ''doomsday'' events ($D_1$ and $D_2$ in the
Figure).

@@

one of the CETIs, are defined as ''contact'' events, shown as $C_1$

@@

CETIs, are denominated ''blackout'' events, shown as $B_1$ and $B_2$

@@

It is important to notice that there is a time delay between the contact
events of the two CETIs involved in this analysis, and also a time
delay between the two blackout events, so that the time range when a
bidirectional contact is possible occurs between the maximum time of
the contact points and the minimum time between the blackout points.

%
% Schematic representation of the growing communicating
%   sphere, over the space--time diagrams, where time is represented on
%   the horizontal direction, and space is represented in the vertical
%   direction.  In (a) the sphere is
%   growing as the surface of first contact has not reached the maximum
%   distance.  In (b) it has reached the maximum distance, so that it
%   remains at the same size.  After a Doomsday event, the signals can
%   still be observed, but the surface of last contact grows.
%   %
%   Schematic representation of two emitters, $E_1$ and $E_2$,
%   that reach each other at different times.  The time span for $E_i$
%   is $(A_i, D_i)$, for $i=\{1,2\}$.  Emitter $i$ can listen to
%   emitter $j$ between $C_{ij}$ and $B_{ij}$.
%   The type and length of causal contact in both directions depend on
%   the distance and time lag between the awakening events, the maximum
%   distance that a signal can reach and the time period in which each
%   emitter is active.
%
%
%
The initially growing communicating sphere, is shown over
space--time diagrams, where time is represented on the horizontal
direction, and space is represented in the vertical direction.
%
In the upper panel (a), the sphere is growing as the surface of first
contact has not reached the maximum distance.
%
In the middle panel (b) it has reached the maximum distance, so that
it remains at the same size.
%
After a Doomsday event (bottom panel, b), the signals can still be
observed, but the surface of last contact grows.
%
We also show in this Figure a schematic representation of two
emitters, E$_1$ and E$_2$, that reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the
distance and time lag between the awakening events, the maximum
distance that a signal can reach and the time period in which each
emitter is active.
%
In the Fig.~\ref{F_messages} we show a schematic representation of the
possible cases in which a CETI A can be in causal contact with a CETI
B which appears later.
%
The duration of the causal contact in one direction depends on several
factors, mainly $\Delta$t~A , $\Delta$t~C , and $\Delta$X.

@@

The temporal structure of the process is defined by two distribution
parameters that represent the mean time interval that an intelligent
civilization can emit and receive signals, and the mean time interval
between the emergence of an intelligent comunicating civilization and
the next one.
%
The spatial structure of the simulation is given by the size and shape
of the Galactic Habitable Zone and the maximum distance a signal can
travel to be detected.
%
The parameters for the temporal distributions also determine the
spatial properties since the density of active cetis in the Galaxy
depend on these two parameters.
%
For example, a small $\tau_a$ and a large $\tau_s$ will produce a
Galaxy densely populated woth CETIs, and viceversa.
%
Also, some hypotheses must be made in order to complete the
simulation.
%
\citet{forgan_spatiotemporal_2011} argues that the times at which
different civilizations become intelligent 
follow a Gaussian distribution, and then the distribution of
inter-arrival times is an inverse exponential.
%
We chose to assume that the distribution of the times of A events is a
stationary Poisson process, and then the distribution of the times between the
appeareance of new consecutive CETIs is exponential.
%
Regarding the duration of a CETI, we propose that the distribution of
the duration of CETIs is a stationary exponential distribution.
%
\citet{balbi_impact_2018} stress the fact that it would be desirable to arrive
at a theoretical statistical distribution of the lifespan of
civilizations, preferentially on the basis of the underlying
astrophysical and biological processes.
%
\citet{maccone_lognormals_2014} argues that this distribution should be a log normal.
(EXPAND)

There is no a temporal window for the model, i.e., the simulation
starts when the process is already stable, and ends before any
galactic effect could modify or affect of fixed variables.
%
The probability for the rise of a CETI is homogeneous over the GHZ.
%
Although the Galaxy has a well known spiral structure, the assumed
sparcity of civilizations makes this assumption very reasonably.
%
Otherwise, if the distribution of civilizations is not sparce, it
could be the case that the spiral arms would host most of the
civilizations, and contacts are frecuent between closely located
civilizations.
%
In such a case we could expect that the interpretation of our results
could be considered as pesimist for the stablishment of a contact with
our planet.
%
We also limit the posibilities of life or other types of civilizations
to the hypothesis usually stated for the definition of the GHZ
\citep{dayal_habitability_2016,gonzalez_galactic_2001,lineweaver_galactic_2004,gonzalez_habitable_2005,morrison_extending_2015,haqq-misra_evolution_2019,rahvar_cosmic_2016,gobat_evolution_2016,
rahvar_cosmic_2016}.
%
\citet{vukotic_astrobiological_2012} propose probabilistic cellular
automata model to explain the astrobiological history of the Milky
Way.
%
This means that we set aside possibel civilizations that could survive
in severe conditions or unstable systems, which would prevent the
appeareance of life as we know it.

\setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{clccc}
\hline
   \multicolumn{2}{l}{independent variables} &min&max&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive awakenings & 4000 y & 200000 y & 50\\ 
   $\tau_{s}$ & Mean lifetime of a CETI                                 & 10000 y & 500000 y& 50\\ 
   D$_{max}$ & Maximum reach of a message  &  \multicolumn{3}{l}{500, 1000, 10000, 20000, 40000, 80000 lyr}  \\
\hline
   \multicolumn{4}{l}{fixed variables and assumptions} &value \\
\hline
   & statistical properties of all CETIs &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & 2--dimensional ring &&\\
   R$_{SFC}$   & Inner radius of the GHZ     &&&20000 ly\\
   R$_{SLC}$   & Outer radius of the GHZ     &&&60000 ly\\
   t$_{max}$ & Time span of the simulation  &&&1.e6 y\\
\hline
   \multicolumn{5}{l}{discrete events}\\
\hline
   A & Awakening: a civilization starts a SETI program  &&&\\
   B & Blackout: the end of the communication chanel stops &&&\\
   C & Contact: a new causal contact is produced &&&\\
   D & Doomsday: a civilization ends its communication capabilities&&&\\
\hline

@@

While the aforementioned assumptions are basic conditions for most of
the stochastic processes observed in nature, we need to make stronger
assumptions related to the nature of the message.
%
The most simple assumption about the message itself, is that it
travels at light speed.
%
With this, we are considering messages sent through, for example,
electromagnetic radiation or even gravitational radiation, but we set
appart messages sent with mechanical means or physical objects, or
through some unknown technology that violates the known so far laws of
physics.
% 
For the communication of CETIs through messages sent isotropically, we
assume that the capacity to emit signals and to receive signals ocurr
at the same time.
%
This means that the ability to find another civilization develops at
the same time that the ability and intention to emit a signal in all
direction, intented to be detected by an unknown civilization.
%
Although there are several reasons to think that this could not be the
case, at large time scales it can be considered that both abilities
occurr roughly at coincident epochs.
%
Another essential assumption is that all CETIs use the same signal
power, so that there is a maximum distance out to which it can be
detected.
%
It is worth mentioning that we are considering in our experimental
setup a system composed by several emmiters and receivers across the
Galaxy, under the hypothesis proposed before.
%
In such a system we compute probabilities of a CETI making contact
with another CETI, and the results must be interpreted as for a
randomly chosen CETI, given certain conditions of homogeneityon an
ensemble of equivalent galaxies.
%
We stress the fact that we are not centered on the probabilities for
the Earth itselft, but on a generalized CETI.
%
Regarding the extent of the signals, we know that the distance from
which a signal from Earth could be detected using the current
technology is of a few hundreds light years, given that the signal was
sent to a specific direction.

@@

The exponential distribution of lifespan and waiting times is
justified by considering that the process of appeareance of life in
the galaxy is homogeneous and stationary.
%
This means that there is no preferred location within the GHZ for the
spontaneos appeareance of life, and that the emergence of a
civilization is independent of the existence of previous civilizations 
in the galaxy.
%            
This is equivalent to proposing a Poisson process for the emergence of
CETIs, since there is a close relation between the number of events in
time or space and the waiting time or separation, respectively.
%
That is, these are two alternative approaches to describing the same
process, a Poisson distribution for the number of events implies an
exponential distribution for their sepations, and viceversa.
%
It should be emphasized that the exponential laws used in this work
are assumed, although instead of analyzing results from a particular
parameter chosen ad--hoc, we explore the hypothesis space and analyze
the impact of the values of these parameters on the results.
%

@@

Given a parametric model $M(\tau_a, \tau_s, D_{max})$, which includes the
functional forms of the statistical distributions and fixed values for the
hypotheses, a discrete event simulation is performed by keeping track of a set
of variables that change each time a meaningful event is produced.
%
The main variables that follow the evolution of the simulation are the positions
of stars, which are sampled randomly within the GHZ, the time of the awakening
of each CETI (A event), and the time of the vanishing of each CETI (D event).
%
The variables that can be deduced from the previous ones include the number of
CETIs in casual contact with at least another CETI at a given time, the number
of CETIs as a function of time, the number of CETIs that receive a message at
least one time, the number of CETIs that receive a message at least one time and
successfully deliver an answer and the number distribution of waiting time to
receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D)
is produced.

@@

\begin{figure} \centering
   %
   \includegraphics[width=0.5\textwidth]{F1.pdf}
   %
   \caption{Empirical cumulative distributions of the number contacts
   for CETIs in four different samples (M5, M6, M7, M8) with long
   range signal reach (D$_{max}$=10000).
   %
   }
   % 
\label{F_number_of_contacts} \end{figure}

@@

\subsection{Model for CETIs}
%{{{

@@

\begin{enumerate}
   \item[(A)] A new CETI appears (Awakening)
   \item[(D)] An old CETI dissappears (Doomsday)
   \item[(C)] A new causal contact is stablished (Contact)
   \item[(B)] An existing causal contact is interrupted (Blackout)
\end{enumerate}

The event of type B is produced when one of the two CETIs halts in its
capability of emmiting and receiving signals.

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{matrix_1.pdf}
   \caption{Rate of CETIs that never make contact (listening) as a
   function of $\tau_a$ and $\tau_s$, for 
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{matrix_2.pdf}
   \caption{Rate of CETIs that listen at the moment of the
   awakening, as a funcion of $\tau_a$ and $\tau_s$, for
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_C_at_A}
\end{figure*}

The system is updated each time an event is produced.
%
The time marks for each event (A and D for each CETI and C and B for each causal
contact) are stored as a result of the simulation.
%
Also, the list of active CETIs is obtained as a function of time.
%
Some fixed variables must be set in order to carry out the simulation: the size
of the Galactic Habitable Zone, the mean lifetime of a CETI, $\tau_s$, the mean
waiting time for the appeareance of another CETI, $\tau_a$, and the maximum
distance a message can be detected.

@@

We implemented the simulation of a regular grid of models varying over
the hypothesis space, which covers 15000 different models.
%
For each model, we simulated 50 realizations with different random
seeds, in order to estimate the variance.
%
The simulated parameters, $\tau_a$, $\tau_s$ cover the ranges
4000-200000 and 10000-500000, respectively, with a partition of 50
values for each parameter.
%
For the D$_{max}$ parameter, we chose the values 500, 1000, 10000,
20000, 40000 and 80000 lyr.
%
This setup makes a total of 750000 simulations, each covering a time
range of one million years.

>>>>

\Abbreviations{MPL: Communicating Extraterrestrial Intelligence, DE:
Discrete Event, GHZ: Galactic Habitable Zone}

@@

Simulations are suitable tools to analyze systems that evolve with
time and involve randomness.
%
An advantadge of simulations compared to theoretical approaches, is
that the former usually require less assumptions and simplifications,
and can be applied to systems where a theoretical model can not be
found.
%
In particular, many complex stochastic processes that can be described
by the evolution of the state of a system, can be efficiently modeled
with the discrete--event (hereafter, DE) simulation approach.
%
A system described with the DE paradigm is characterized by a set of
actors and events, where actors interact causally through a series of
events on a timeline and process these events in chronological order
\citep{ptolemaeus_system_2014, chung_simulation_2003,
ross_simulation_2012}.
%
Each event produces changes in the values of the variables that define
the process, and thus the corresponding change in the state of the
simulated system.
%
This method is well suited for the particular case of the difusion of
intelligent signals in the Galaxy, and allows to explore several
models easily.
%
We simulate the statistical properties of a set of points in space and
time that have the ability of emitting and communicating messages at
light speed, hereafter dubbed ''message processing locations'' (MPL).
%
These locations have the capability of emitting and receiving messages
in all directions.
%
It is worth mentioning that this is a general approach, and not
necessarily a MPL is the host of an intelligent civilization.
%
It can be associated with a set of planets where life has developed,
became intelligent, reached the skills required to find the right
communication channel, sustain a search and stablish a contact.
%
Alternative message processing entities could be considered, for
example interstellar beacons where life has ceased to exist but
continue with its emission or communication stations stablished by
probes or left intentionally by intelligent beings \citep[see, e.g.,
][]{peters_outer_2018, barlow_galactic_2013}.
%
In principle, these strategies could affect our results since it would
be easier to configure a cluster of MPLs that spread in time.
%
However, we do not consider this highly speculative alternatives at
this point.
%
% have been proposed
%
For the purpose of this analysis, only the communication capability is
relevant, since we study the causal contacts between the locations.
%
% communicating, civilization, intelligent, message, processing,
% handling, locations CALL, MPI, MPL, MPL
%
The system is defined by a number of actors that represent MPLs and
appear at different instants in time, generating events that produce
meaningful changes in the variables that describe the system, i.e., in
the arrangement of MPLs and their network of causal contacts.
%
For example, the appeareance of a new MPL in a region which is filled
with a signal emitted by another MPL, will increase the number of
active MPLs and the number of pairs of MPLs in causal contact.
%
Assuming some simple hypotheses, the discrete events method can be
performed taking into account a small number of variables, which allow
to analyze the variation of the results in the model parameter space.

In what follows, we describe the experimental setup chosen to estimate
the probabilities of causal contacts and several derived quantities in
terms of three independent parameters, namely, the mean time span
between the appeareance of consecutive MPLs, ($<\tau_a>$), their mean
lifetime ($<\tau_s>$), and the maximum distance a signal can be
detected by another MPL ($D_{max}$).
%
Intuitively, the shorter the $\tau_a$ parameter and the larger the
$\tau_s$ and $D_{max}$ parameters are, the greater the probability of
the existence of causal contacts between pairs of MPLs would be.
%
We also need to propose theoretical distribution functions for both
the distribution of lifespans ($\tau_s$) and the distribution of the
number of MPLs per unit time \citep{maccone_evolution_2014,
Sotos_biotechnology_2019}, related to the time span between the
appeareance of consecutive MPLs (since when $<\tau_a>$ is shorter, it
produces a greater density of MPLs).
%
The shapes of these distributions are set to a fixed law, as discussed
in Sec.~\ref{SS_PDF_shape}.

@@

involved locations being concurrent or active at the same time.

@@

Although a MPL could be active for enough time so as to transmit a
message at large distances, the limited power of the message and the
dilution that runs as the squared distance from the source, imposes a
detectability limit.

@@

communicating capacity, each MPL will fill a spherical shell region of
the Galaxy, limited by two concentric spherical surfaces.

@@

central MPL until it reaches the maximum distance $D_{max}$.

@@

Thus, the volume reached causally by a MPL is initially a growing

@@

the central MPL, with a delay with respect to the SFC equivalent to
the lifetime of the MPL, and produces a spherical shell region.

@@

Any other MPL within this region will be aware of the originating MPL,
even if it has disapeared before the time of contact.

@@

until the surface of last contact reaches $D_{max}$, producing as a
result the loss of all signal from the central MPL.

@@

disk is negligigle with respect to the radius of the disk, so that the
MPLs are placed on a 2D plane disk.

@@

of the region causally connected to the central MPL.

@@

At the end of the lifetime of a MPL, a growing surface of last

@@

In this scheme, the entire lifetime of a MPL can then be represented

@@

that result from the existence and communications of two MPLs (panel
(b)).

@@

The points in time where the MPLs aquire their communicating capacity
(CC) are dubbed ''awakening'' events ($A_1$ and $A_2$ in the Figure).

@@

Simmilarly, the points in time where the MPLs loss their
communicating capacity are dubbed ''doomsday'' events ($D_1$ and $D_2$
in the Figure).

@@

one of the MPLs, are defined as ''contact'' events, shown as $C_1$

@@

MPLs, are denominated ''blackout'' events, shown as $B_1$ and $B_2$

@@

It is important to notice that there is a time delay between the
contact events of the two MPLs involved in this analysis, and also a
time delay between the two blackout events, so that the time range
when a bidirectional contact is possible occurs between the maximum
time of the contact points and the minimum time between the blackout
points.

%
% Schematic representation of the growing communicating
%   sphere, over the space--time diagrams, where time is represented on
%   the horizontal direction, and space is represented in the vertical
%   direction.  In (a) the sphere is
%   growing as the surface of first contact has not reached the maximum
%   distance.  In (b) it has reached the maximum distance, so that it
%   remains at the same size.  After a Doomsday event, the signals can
%   still be observed, but the surface of last contact grows.
%   %
%   Schematic representation of two emitters, $E_1$ and $E_2$,
%   that reach each other at different times.  The time span for $E_i$
%   is $(A_i, D_i)$, for $i=\{1,2\}$.  Emitter $i$ can listen to
%   emitter $j$ between $C_{ij}$ and $B_{ij}$.
%   The type and length of causal contact in both directions depend on
%   the distance and time lag between the awakening events, the maximum
%   distance that a signal can reach and the time period in which each
%   emitter is active.
%
%
%
The initially growing communicating sphere, is shown over
space--time diagrams, where time is represented on the horizontal
direction, and space is represented in the vertical direction.
%
In the upper panel (a), the sphere is growing as the surface of first
contact has not reached the maximum distance.
%
In the middle panel (b) it has reached the maximum distance, so that
it remains at the same size.
%
After a Doomsday event (bottom panel, b), the signals can still be
observed, but the surface of last contact grows.
%
We also show in this Figure a schematic representation of two
emitters, E$_1$ and E$_2$, that reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the
distance and time lag between the awakening events, the maximum
distance that a signal can reach and the time period in which each
emitter is active.
%
In the Fig.~\ref{F_messages} we show a schematic representation of the
possible cases in which a MPL A can be in causal contact with a MPL
B which appears later.
%
The duration of the causal contact in one direction depends on several
factors, mainly $\Delta$t~A , $\Delta$t~C , and $\Delta$X.

@@

The temporal structure of the process is defined by two distribution
parameters that represent the mean time interval that an intelligent
civilization can emit and receive signals, and the mean time interval
between the emergence of an intelligent comunicating civilization and
the next one.
%
The spatial structure of the simulation is given by the size and shape
of the Galactic Habitable Zone and the maximum distance a signal can
travel to be detected.
%
The parameters for the temporal distributions also determine the
spatial properties since the density of active cetis in the Galaxy
depend on these two parameters.
%
For example, a small $\tau_a$ and a large $\tau_s$ will produce a
Galaxy densely populated with MPLs, and viceversa.
%
Also, some hypotheses must be made in order to complete the
simulation.
%
\citet{forgan_spatiotemporal_2011} argues that the times at which
different civilizations become intelligent 
follow a Gaussian distribution, and then the distribution of
inter-arrival times is an inverse exponential.
%
We chose to assume that the distribution of the times of A events is a
stationary Poisson process, and then the distribution of the times between the
appeareance of new consecutive MPLs is exponential.
%
Regarding the duration of a MPL, we propose that the distribution of
the duration of MPLs is a stationary exponential distribution.
%
\citet{balbi_impact_2018} stress the fact that it would be desirable to arrive
at a theoretical statistical distribution of the lifespan of
civilizations, preferentially on the basis of the underlying
astrophysical and biological processes.
%
\citet{maccone_lognormals_2014} argues that this distribution should be a log normal.
(EXPAND)

There is no a temporal window for the model, i.e., the simulation
starts when the process is already stable, and ends before any
galactic effect could modify or affect of fixed variables.
%
The probability for the rise of a MPL is homogeneous over the GHZ.
%
Although the Galaxy has a well known spiral structure, the assumed
sparcity of civilizations makes this assumption very reasonably.
%
Otherwise, if the distribution of civilizations is not sparce, it
could be the case that the spiral arms would host most of the
civilizations, and contacts are frecuent between closely located
civilizations.
%
In such a case we could expect that the interpretation of our results
could be considered as pesimist for the stablishment of a contact with
our planet.
%
We also limit the posibilities of life or other types of civilizations
to the hypothesis usually stated for the definition of the GHZ
\citep{dayal_habitability_2016,gonzalez_galactic_2001,lineweaver_galactic_2004,gonzalez_habitable_2005,morrison_extending_2015,haqq-misra_evolution_2019,rahvar_cosmic_2016,gobat_evolution_2016,
rahvar_cosmic_2016}.
%
\citet{vukotic_astrobiological_2012} propose probabilistic cellular
automata model to explain the astrobiological history of the Milky
Way.
%
This means that we set aside possible civilizations that could survive
in severe conditions or unstable systems, which would prevent the
appeareance of life as we know it.

\setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{clccc}
\hline
   \multicolumn{2}{l}{independent variables} &min&max&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive awakenings & 4000 y & 200000 y & 50\\ 
   $\tau_{s}$ & Mean lifetime of a MPL                                 & 10000 y & 500000 y& 50\\ 
   D$_{max}$ & Maximum reach of a message  &  \multicolumn{3}{l}{500, 1000, 10000, 20000, 40000, 80000 lyr}  \\
\hline
   \multicolumn{4}{l}{fixed variables and assumptions} &value \\
\hline
   & statistical properties of all MPLs &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & 2--dimensional ring &&\\
   R$_{SFC}$   & Inner radius of the GHZ     &&&20000 ly\\
   R$_{SLC}$   & Outer radius of the GHZ     &&&60000 ly\\
   t$_{max}$ & Time span of the simulation  &&&1.e6 y\\
\hline
   \multicolumn{5}{l}{discrete events}\\
\hline
   A & Awakening: a civilization starts a SETI program  &&&\\
   B & Blackout: the end of the communication chanel stops &&&\\
   C & Contact: a new causal contact is produced &&&\\
   D & Doomsday: a civilization ends its communication capabilities&&&\\
\hline

@@

While the aforementioned assumptions are basic conditions for most of
the stochastic processes observed in nature, we need to make stronger
assumptions related to the nature of the message.
%
The most simple assumption about the message itself, is that it
travels at light speed.
%
With this, we are considering messages sent through, for example,
electromagnetic radiation or even gravitational radiation, but we set
appart messages sent with mechanical means or physical objects, or
through some unknown technology that violates the known so far laws of
physics.
% 
For the communication of MPLs through messages sent isotropically, we
assume that the capacity to emit signals and to receive signals ocurr
at the same time.
%
This means that the ability to find another civilization develops at
the same time that the ability and intention to emit a signal in all
direction, intented to be detected by an unknown civilization.
%
Although there are several reasons to think that this could not be the
case, at large time scales it can be considered that both abilities
occurr roughly at coincident epochs.
%
Another essential assumption is that all MPLs use the same signal
power, so that there is a maximum distance out to which it can be
detected.
%
It is worth mentioning that we are considering in our experimental
setup a system composed by several emmiters and receivers across the
Galaxy, under the hypothesis proposed before.
%
In such a system we compute probabilities of a MPL making contact
with another MPL, and the results must be interpreted as for a
randomly chosen MPL, given certain conditions of homogeneityon an
ensemble of equivalent galaxies.
%
We stress the fact that we are not centered on the probabilities for
the Earth itselft, but on a generalized MPL.
%
Regarding the extent of the signals, we know that the distance from
which a signal from Earth could be detected using the current
technology is of a few hundreds light years, given that the signal was
sent to a specific direction.

@@

The exponential distribution of lifespan and waiting times is
justified by considering that the process of appeareance of life in
the galaxy is homogeneous and stationary.
%
This means that there is no preferred location within the GHZ for the
spontaneos appeareance of life, and that the emergence of a
civilization is independent of the existence of previous civilizations 
in the galaxy.
%            
This is equivalent to proposing a Poisson process for the emergence of
MPLs, since there is a close relation between the number of events in
time or space and the waiting time or separation, respectively.
%
That is, these are two alternative approaches to describing the same
process, a Poisson distribution for the number of events implies an
exponential distribution for their sepations, and viceversa.
%
It should be emphasized that the exponential laws used in this work
are assumed, although instead of analyzing results from a particular
parameter chosen ad--hoc, we explore the hypothesis space and analyze
the impact of the values of these parameters on the results.
%

@@

Given a parametric model $M(\tau_a, \tau_s, D_{max})$, which includes the
functional forms of the statistical distributions and fixed values for the
hypotheses, a discrete event simulation is performed by keeping track of a set
of variables that change each time a meaningful event is produced.
%
The main variables that follow the evolution of the simulation are the positions
of stars, which are sampled randomly within the GHZ, the time of the awakening
of each MPL (A event), and the time of the vanishing of each MPL (D event).
%
The variables that can be deduced from the previous ones include the number of
MPLs in casual contact with at least another MPL at a given time, the number
of MPLs as a function of time, the number of MPLs that receive a message at
least one time, the number of MPLs that receive a message at least one time and
successfully deliver an answer and the number distribution of waiting time to
receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D)
is produced.

@@

\begin{figure} \centering
   %
   \includegraphics[width=0.5\textwidth]{F1.pdf}
   %
   \caption{Empirical cumulative distributions of the number contacts
   for MPLs in four different samples (M5, M6, M7, M8) with long
   range signal reach (D$_{max}$=10000).
   %
   }
   % 
\label{F_number_of_contacts} \end{figure}

@@

\subsection{Model for MPLs}
%{{{

@@

\begin{enumerate}
   \item[(A)] A new MPL appears (Awakening)
   \item[(D)] An old MPL dissappears (Doomsday)
   \item[(C)] A new causal contact is stablished (Contact)
   \item[(B)] An existing causal contact is interrupted (Blackout)
\end{enumerate}

The event of type B is produced when one of the two MPLs halts in its
capability of emmiting and receiving signals.

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{matrix_1.pdf}
   \caption{Rate of MPLs that never make contact (listening) as a
   function of $\tau_a$ and $\tau_s$, for 
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{matrix_2.pdf}
   \caption{Rate of MPLs that listen at the moment of the
   awakening, as a funcion of $\tau_a$ and $\tau_s$, for
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_C_at_A}
\end{figure*}

The system is updated each time an event is produced.
%
The time marks for each event (A and D for each MPL and C and B for each causal
contact) are stored as a result of the simulation.
%
Also, the list of active MPLs is obtained as a function of time.
%
Some fixed variables must be set in order to carry out the simulation: the size
of the Galactic Habitable Zone, the mean lifetime of a MPL, $\tau_s$, the mean
waiting time for the appeareance of another MPL, $\tau_a$, and the maximum
distance a message can be detected.

@@

We implemented the simulation of a regular grid of models varying over
the hypothesis space, which covers 15000 different models.
%
For each model, we simulated 50 realizations with different random
seeds, in order to estimate the variance.
%
The simulated parameters, $\tau_a$, $\tau_s$ cover the ranges
4000-200000 and 10000-500000, respectively, with a partition of 50
values for each parameter.
%
For the D$_{max}$ parameter, we chose the values 500, 1000, 10000,
20000, 40000 and 80000 lyr.
%
This setup makes a total of 750000 simulations, each covering a time
range of one million years.
%
In Table \ref{T_simu_hypotheses} we show the range of values of these parameters explored 
in our numerical experiments.

As a product of the simulations, several quantities are obtained.
%
Some quantities are directly derived from the discrete events, namely,
the ID of emitting MPL, the ID of emitting (receiving) MPL, the
position in the galaxy, the time of Awakening (Contact) and the time
of Doomsday  (Blackout) for each MPL.
%
This structure is always for the A-D cycle of each emitting MPL, and
repeats for each contact (C-B).
%
We can also derive the additional quantities.
%
The age of a MPL is the time elapsed from A to D.
%
%
%given time, ceti\_e is the MPL emitting signals (emiter), ceti\_r is
%the MPL receiving signals (receiver), ceti\_c is the ceti that
%listen at least another ceti (citizen) and ceti\_h is the MPL ceti
%that is lestened by at least another MPL.
%
The time span of a MPL listening another or being listened by another
can also be easily derived from the results of a realization.
%
This way we can also compute the distribution in the galaxy of cetis
that reach contact and the waiting time until the first contact or the
waiting time until the next contact.
%
Regarding the properties of the population of MPLs and its evolution,
we analyze the fraction of awaken time a ceti is listening at least
another ceti, the age of contacted MPLs at first contact, the
distribution of time to wait until next contact, the fraction of cetis
where the first contact is given at the awakening, the distribution of
the number of contacts for each MPL, the distribution of the maximum
number of contacts for each MPL, the distribution of the number of
contacts as a function of MPL age, the number of contacts as a
function of time in the galaxy, the rate of MPLs that succeed in
contact, and the distribution of distances between contacted MPLs.
%
Usefull derived quantity is the duration of two-way communication
channels the or the fraction of contacts that admit a response.
% 
We also analyze the relations between the distance to ceti vs. time of
double communication, the distance to ceti vs. age of contacted ceti,
the age of a ceti and the maximum number of contacted cetis before D,
or the lifespan of a ceti vs. max number of contacts.
%
All these quantities can be analyzed as a function of the simulation
parameters.
%
We chose eight models that are on fairly oposed regions of the
explored parameter space, and cover short/long lifetimes, short/long
waiting times for the next MPL to appear in the Galaxy and
short/large range reach of the signal, including all possible
combinations.
%
The details of each model are summarized in the
Table~\ref{T_selected_models}.
